---
title: "D is for Demography"
author: "Monica Alexander"
geometry: paperheight=4.6in,paperwidth=5.7in,margin=1in
mainfont: Helvetica
header-includes:
  - \pagenumbering{gobble} 
output:
    pdf_document:
      latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev="cairo_pdf", message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(LexisPlotR)
library(readxl)
library(here)
library(janitor)
library(kableExtra)
```


\centering

# A is for Age
```{r}
lexis_grid(year_start = 1970, year_end = 2020, age_start = 0, age_end = 50, delta = 5) %>%
  lexis_age(age = 30, delta = 5)
```



# B is for Births

```{r}
births <- read_xlsx("data/WPP2019_FERT_F01_BIRTHS_BOTH_SEXES.xlsx", skip = 16)
births <- births %>% 
  select(`Region, subregion, country or area *`, `1950-1955`:`2015-2020`) %>% 
  rename(region = `Region, subregion, country or area *`) %>% 
  pivot_longer(-region, names_to = "period", values_to = "births") %>% 
  mutate(births = as.numeric(births)) %>% 
  mutate(year = as.numeric(str_sub(period, 1, 4))+2.5)

births %>% 
  filter(region %in% c("Australia", "Canada")) %>% 
  ggplot(aes(year, births, color = region)) + 
  geom_point() + geom_line(lwd = 1.3) + 
  theme_bw(base_size = 14) + 
  #ggtitle("Births in Australia and Canada") + 
  scale_color_brewer(palette = "Set1", name = "country")+
  labs(caption = "Data from WPP 2019")
```


# C is for Cohort

```{r}
lexis_grid(year_start = 1970, year_end = 2020, age_start = 0, age_end = 50, delta = 5) %>%
  lexis_cohort(cohort = 1970, delta = 5)
```


# D is for Demography!

\includegraphics{fig/demogseal2.png}


\newpage

# E is for Euler-Lotka equation

\vspace{10mm}

\Huge
$$
1 = \int e^{-ra}l(a)f(a)da
$$

# F is for Fertility

```{r}
pop_female <- readxl::read_xlsx(here("data/WPP2019_POP_F07_3_POPULATION_BY_AGE_FEMALE.xlsx"), sheet = 1, skip = 16)
pop_male <- readxl::read_xlsx(here("data/WPP2019_POP_F07_2_POPULATION_BY_AGE_MALE.xlsx"), sheet = 1, skip = 16)
pop_female$sex <- "F"
pop_male$sex <- "M"
pop <- bind_rows(pop_female, pop_male)

rm(pop_female, pop_male) # remove the sex-specific copies

# clean up a little and make into a long tibble. we need to make the last age group 95+ to match with deaths

pop <- pop %>% 
  rename(location = `Region, subregion, country or area *`,
         year = `Reference date (as of 1 July)`) %>% 
  select(location, year,sex,  `0-4`:`100+`) %>% 
  mutate_at(.vars = vars(`0-4`:`100+`), .funs = funs(as.numeric(.))) %>% 
  mutate(`95+` = `95-99`+ `100+`) %>% 
  select(-(`95-99`:`100+`)) %>% 
  pivot_longer(`0-4`:`95+`, names_to = "age_group", values_to = "population")

```


```{r}
age_groups <- unique(pop$age_group) # get all the unique age groups. these are in the order we want

pop <- pop %>% 
  mutate(age_group = factor(age_group, levels = age_groups))

pop %>% 
  filter(location %in% c("Japan", "Nigeria"), year %in% c(2015)) %>%
  group_by(location, year) %>% 
  mutate(location = ifelse(location == "Nigeria", "Nigeria (high fertility)", "Japan (low fertility)")) %>% 
  mutate(proportion = population/sum(population)) %>% 
  mutate(proportion=ifelse(sex=="M", -proportion, proportion)) %>% 
ggplot(aes(x = age_group, y = proportion, fill = sex)) + 
  facet_grid(.~location)+
  geom_bar(stat="identity")+
  ylab("Proportion of population")+
  xlab("Age group") +
  coord_flip() + 
  scale_y_continuous(breaks = seq(-.1, .1, 0.05), 
                     labels = c(seq(.1, 0, -0.05), seq(0.05, .1, 0.05))) + 
  scale_fill_brewer(palette = "Set1") +
  theme_bw(base_size = 14) 
```

# G is for Gompertz mortality

```{r}
d <- read_table(here("data/AUS_Mx_1x1.txt"), skip = 2, col_types = 'dcddd')

# tidy up 
d <- d %>% 
  clean_names() %>% 
  mutate(age = as.numeric(ifelse(age=="110+", "110", age))) %>% 
  filter(age<101) %>% # older ages problematic in earlier years
  pivot_longer(female:total, names_to = "sex", values_to = "mx")

d_gomp <- d %>% 
  filter(age>29) %>% 
  mutate(log_mx = log(mx)) %>% 
  group_by(year, sex) %>% 
  summarise(alpha = exp(coef(lm(log_mx~age))[1]),
            beta = coef(lm(log_mx~age))[2])
d_gomp %>% 
  pivot_longer(alpha:beta, names_to = "parameter", values_to = "value") %>% 
  filter(sex!='total') %>% 
  ggplot(aes(year, value, color = sex)) + geom_point() + 
  facet_wrap(~parameter, scales = "free_y") + 
  scale_color_brewer(palette = "Set1") + 
  theme_bw(base_size = 14)+
  labs(title = "Gompertz parameters for Australian males", caption = "Data from HMD")
```


\newpage
# H is for Hutterites


\begin{table}[h!]
\centering
\begin{tabular}{l|r}
Age & $n(x)$ \\ \hline
15  & 0.36   \\
20  & 0.46   \\
25  & 0.431  \\
30  & 0.396  \\
35  & 0.321  \\
40  & 0.167  \\
45  & 0.024 
\end{tabular}
\end{table}

\tiny
Data reproduced from Wachter 2014

# I is for Immigration
\vspace{-2mm}
\includegraphics[width = 0.8\textwidth]{fig/pr_state.png}

# J is for 'J-shape' mortality


```{r}
d <- read_table(here("data/AUS_Mx_1x1.txt"), skip = 2, col_types = 'dcddd')

# tidy up 
d <- d %>% 
  clean_names() %>% 
  mutate(age = as.numeric(ifelse(age=="110+", "110", age))) %>% 
  filter(age<101) %>% # older ages problematic in earlier years
  pivot_longer(female:total, names_to = "sex", values_to = "mx")

d %>% 
  filter(year %in% seq(1955, 2015, by = 10), sex=="male") %>% 
  ggplot(aes(age, mx, color = factor(year))) + 
  geom_line(lwd = 1.3) + 
  scale_color_viridis_d(name = "year")+
  scale_y_log10() + 
  labs(title = "Age-specific mortality, Males, Australia", caption = "Data from HMD") + 
  theme_bw(base_size = 14)
```

# K is for Keyfitz momentum

```{r}
df <- read_xlsx(here("data", "WPP2019_FERT_F07_AGE_SPECIFIC_FERTILITY.xlsx"), skip = 16)

df <- df %>% 
  rename(region = `Region, subregion, country or area *`, period = Period) %>% 
  select(region, period, `15-19`:`45-49`) %>% 
  pivot_longer(-(region:period), names_to = "age", values_to = "Fx") %>% 
  mutate(year = as.numeric(substr(period, 1, 4)), 
         age = as.numeric(substr(age, 1, 2)),
         Fx = as.numeric(Fx)) %>% 
  mutate(age = as.numeric(age), Fx = Fx/1000)

dl <- read_xlsx(here("data", "WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx"), skip = 16)

dl <- dl %>% 
  rename(region = `Region, subregion, country or area *`,
         Lx = `Number of person-years lived L(x,n)`,
         age = `Age (x)`,
         period = Period) %>% 
  select(region, period, age, Lx) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)), Lx = as.numeric(Lx)/10^5)

nLx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada", age<85) %>% 
  select(Lx) %>% pull()

## need to fix first age group

nLx <- c(sum(nLx[1:2]), nLx[3:length(nLx)])

nFx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(Fx = ifelse(is.na(Fx), 0, Fx)) %>% 
  select(Fx) %>% pull()
nFx <- nFx[-1]

NRR <- sum(nFx*nLx*0.4886)
de <- read_xlsx(here("data", "WPP2019_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.xlsx"), skip = 16)

ex <- de %>% 
  rename(region = `Region, subregion, country or area *`,
         ex = `Expectation of life e(x)`,
         age = `Age (x)`,
         period = Period) %>% 
  select(region, period, age, ex) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)), ex = as.numeric(ex)) %>% 
  filter(region == "Canada", age==0, year==2010) %>% 
  select(ex) %>% 
  pull()



d_female <- read_xlsx(here("data", "WPP2019_POP_F07_3_POPULATION_BY_AGE_FEMALE.xlsx"), skip = 16)

Kt <- d_female %>% 
  rename(region = `Region, subregion, country or area *`,
         year = `Reference date (as of 1 July)`) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  filter(region=="Canada", year==2010) %>% 
  pivot_longer(`0-4`:`100+`, names_to = "age", values_to = "pop") %>% 
  rowwise() %>% 
  mutate(age = as.numeric(str_split(age, "-")[[1]][1])) %>% 
  filter(age<85) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  select(pop) %>% 
  pull()


leslie <- function(nLx,
                   nFx, 
                   n_age_groups=17,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  L[1,ncol(L)] <- 0
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}

A <- leslie(nLx, nFx)
eigen_A <- eigen(A)
r_A <- log(Re(eigen_A$value[1]))/5 
b_pre <- 1/sum(exp(-r_A*seq(0, 80, by = 5))*nLx)
NRR <- sum(nFx*nLx*0.4886)
#ex*b_pre/sqrt(NRR)

B <- rbind(A[1,]/NRR, A[-1,])

age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- B%*%K[,i-1] 
}

Kdf <- as_tibble(K)
colnames(Kdf) <- seq(from = 2010, to = (2010+n_projections*5), by = 5)
Kdf <- cbind(age = seq(from = 0, to = 80, by = 5), Kdf)

# get in long format and then add proportion of population in each age group
dk <- Kdf %>% 
  gather(year, population, -age) %>%
  mutate(year = as.numeric(year)) %>% 
  group_by(year) %>%
  mutate(proportion = population/sum(population))

dk %>% 
  group_by(year) %>% 
  summarise(pop = sum(population)) %>% 
  ggplot(aes(year, pop)) + 
  geom_line(lwd = 1.3) + 
  labs(title = "Population over time if NRR went to 1 in 2010", subtitle = "Canada 2010-2250", caption = "Data from WPP 2019") + ylab("population") + theme_bw(base_size = 14)

```

# L is for Life expectancy

\includegraphics{fig/oeppen_vaupel.jpg}
\tiny
From Oeppen and Vaupel (2002)

# M is for Mortality

\includegraphics{fig/opioid_race.png}
\tiny
From https://github.com/MJAlexander/opioid-mcd


# N is for Net Reproduction Ratio

```{r}
ffab <- 0.4886

df %>% 
  left_join(dl) %>% 
  mutate(prod = Fx*Lx*ffab) %>% 
  group_by(region, year) %>% 
  summarise(NRR = sum(prod)) %>% 
  filter(region %in% c("Peru", "Pakistan", "Mozambique", "Canada", "Australia")) %>% 
  ggplot(aes(year, NRR, color = region)) +
  geom_line(lwd = 1.3) + 
  theme_bw(base_size = 14)+
  scale_color_viridis_d(name = "Country")
```

# O is for Old-age dependency ratio

```{r}
pop %>% 
  filter(location %in% c("Australia", "Indonesia")) %>% 
  rowwise() %>% 
  mutate(age = as.numeric(str_split(age_group, "-")[[1]][1])) %>% 
  filter(age>10) %>% 
  mutate(old = ifelse(age>60, 1, 0)) %>% 
  group_by(location, year) %>% 
  summarize(OADR = sum(population[old==1])/sum(population[old==0])) %>%  
  ggplot(aes(year, OADR, color = location)) + geom_line(lwd = 1.3) + 
  theme_bw(base_size = 14) + 
  scale_color_brewer(palette = "Set1", name = "Country")+
  labs(caption = "Data from WPP 2019")
```


# P is for Parity 

```{r}
ppr <- read_table(here("data", "CANpprVHbo.txt"), skip = 2)
ppr <- ppr %>% 
  gather(measure, ratio, -Cohort) %>% 
  mutate(parity =  as.numeric(substr(measure, 6, 6))) %>% 
  rename(cohort = Cohort) %>% 
  select(-measure)
ppr %>% 
  mutate(parity = factor(parity)) %>% 
  ggplot(aes(cohort, ratio, color = parity)) + 
  geom_line(lwd = 1.1) + 
  scale_color_viridis_d() + 
  ylab("PPR") + 
  theme_gray(base_size = 14) + 
  labs(title = "Parity Progression Ratios in Canada", caption = "Data from HFD") +
  theme_bw(base_size = 14)
```


# Q is for Quantum (and tempo)

```{r}
dfc <- read_table(here("data", "CANtfrRRbo.txt"), skip = 2)
dfc_long <- dfc %>% 
  select(Year:TFR5p) %>% 
  select(-TFR) %>% 
  gather(measure, value, - Year) %>% 
  mutate(parity = as.numeric(substr(measure, nchar(measure) , nchar(measure)))) %>% 
  mutate(parity = ifelse(is.na(parity), 5, parity)) %>% 
  select(-measure) %>% 
  rename(tfr = value)

dmac <- read_table(here("data", "CANmabRRbo.txt"), skip = 2)

mab_tfr <- dmac %>% 
  select(Year:MAB5p) %>% 
  select(-MAB) %>% 
  gather(measure, value, - Year) %>% 
  mutate(parity = as.numeric(substr(measure, nchar(measure) , nchar(measure)))) %>% 
  mutate(parity = ifelse(is.na(parity), 5, parity)) %>% 
  select(-measure) %>% 
  rename(mab = value) %>% 
  left_join(dfc_long) 

tfr_adj <- mab_tfr %>% 
  mutate(diff_mab = (mab - lag(mab))/2,
         tfr_dash = tfr/(1-diff_mab)) %>% 
  group_by(Year) %>% 
  summarise(unadjusted = sum(tfr),
            tempo_adjusted = sum(tfr_dash)) 

tfr_adj %>% 
  gather(tfr, value, -Year) %>% 
  ggplot(aes(Year, value, color = tfr)) + 
  geom_line(lwd = 1.1) + 
  labs(title = "TFR and tempo-adjusted TFR", subtitle = "Canada 1944-2011", caption = "Data from HFD") + 
  theme_bw(base_size = 14)
```

# R is for r (the population growth rate)

```{r}
pop %>% 
  filter(location %in% c("Nigeria", "Japan"), sex == "F") %>% 
  group_by(year, location) %>% 
  summarize(pop = sum(population)) %>% 
  arrange(location, year) %>% 
  group_by(location) %>% 
  mutate(growth_rate = ((lead(pop) - pop)/pop)^1/5) %>% 
  ggplot(aes(year, growth_rate, color = location)) + geom_line(lwd = 1.3)+
  theme_bw(base_size = 14) +
  scale_color_brewer(palette = "Set1", name = "Country") + 
  ylab("growth rate")
```


# S is for Survivorship

```{r}
lt <- read_table("http://www.prdh.umontreal.ca/BDLC/data/ont/fltper_5x5.txt", skip = 2)
lt <- lt %>% 
  filter(Year=="2015-2016") %>% 
  mutate(x = c(0,1,seq(5, 110, by = 5)),
         n = lead(x, default = Inf)-x)
lt %>% 
  mutate(lx = lx/100000) %>% 
  ggplot(aes(x, lx)) +
  geom_line(lwd = 1.3) + 
  xlab("age") + 
  theme_bw(base_size = 14) + 
  labs(title = "Survivorship for Ontario, 2015", caption = "data from the Canadian Human Mortality Database")
```


# T is for Total Fertility Rate

```{r}

df <- read_xlsx(here("data", "WPP2019_FERT_F07_AGE_SPECIFIC_FERTILITY.xlsx"), skip = 16)

df <- df %>% 
  rename(region = `Region, subregion, country or area *`, period = Period) %>% 
  select(region, period, `15-19`:`45-49`) %>% 
  pivot_longer(-(region:period), names_to = "age", values_to = "Fx") %>% 
  mutate(year = as.numeric(substr(period, 1, 4)), 
         age = as.numeric(substr(age, 1, 2)),
         Fx = as.numeric(Fx)) %>% 
  mutate(age = as.numeric(age), Fx = Fx/1000)
df %>% 
  group_by(region, year) %>% 
  summarise(tfr = 5*sum(Fx)) %>% 
  filter(region %in% c("Australia", "Nigeria", "Japan", "Indonesia", "India", "Kenya")) %>% 
  ggplot(aes(year, tfr, color = region))  + 
  geom_line(lwd = 1.1) +
  scale_color_viridis_d()+
  geom_hline(yintercept = 2.1) + 
  theme_gray(base_size = 14) + 
  labs(title = "Total fertility rate for selected countries, 1950-2015", caption = "Data from WPP 2019") + 
  theme_bw(base_size = 14)
```


# U is for Uncertainty

\vspace{10mm}
\includegraphics[width = 1\textwidth]{fig/mortality_uncert.pdf}
From Alexander, Zagheni and Barberi (2017)

\newpage
# V is for Vital registration systems

\includegraphics[width = 1.2\textwidth]{fig/crvs.png}

# W is for Weighted mean

```{r}
dmac %>% 
  ggplot(aes(Year, MAB)) + geom_line(lwd = 1.3) + 
  theme_bw(base_size = 14) + 
  labs(title = "Mean age of childbearing, Canada", caption = "Data from HFD") + ylab("age")
```


# X is for $x$ (the age variable)

\footnotesize
```{r}
lt %>%
  filter(x<20) %>% 
  select(x,n, lx, dx, ax, Lx, Tx, ex) %>% kable(booktabs = T,caption = 'Ontario life table (up to age 15), 2015 (Source: CHMD)') %>% 
  kable_styling(latex_options = "hold_position")
```


# Y is for Year (period)

```{r}
lexis_grid(year_start = 1970, year_end = 2020, age_start = 0, age_end = 50, delta = 5) %>%
  lexis_year(year = 1990, delta = 5, fill = "lightblue")
```

# Z is for Zero population growth

```{r}
pop %>% 
  filter(location == "China", sex == "F") %>% 
  group_by(year) %>% 
  summarize(pop = sum(population)) %>% 
  mutate(growth_rate = ((lead(pop) - pop)/pop)^1/5) %>% 
  ggplot(aes(year, growth_rate)) + geom_line(lwd = 1.3)+
  theme_bw(base_size = 14) + 
  labs(title = "Annual growth rate, China female population", caption = "Data from WPP 2019") + ylab("growth rate")
```

